---
title: "Celiac Disease Lab Data — EDA + Modeling"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.path = "figures/", dpi = 100
)

library(tidyverse)   # data manipulation and plotting
library(corrplot)    # correlation matrix visualisation
library(C50)         # C5.0 decision trees
library(gmodels)     # CrossTable
library(class)       # knn
library(tidymodels)  # random forest via parsnip/ranger, yardstick metrics
library(vip)         # variable importance plots

set.seed(1234)
```

## Project Overview

The goal of this project is to explore and model laboratory and clinical data related to celiac disease.
Using a Kaggle dataset, the following steps were performed:

- Exploratory data analysis (EDA)
- Feature engineering and data cleaning
- Dimensionality reduction (PCA)
- Supervised learning using:
  - Decision trees (C5.0)
  - k-nearest neighbors (KNN)
  - Random forest

A key clinical motivation is minimizing **false negatives**, since missing a true celiac diagnosis may lead to
delayed treatment and long-term complications.

All three models are trained and scored on the **same train/test split** and the **same predictors**, so the
comparison at the end reflects model choice rather than differences in setup. Hyperparameters are chosen by
repeated cross-validation on the training set, the test set is evaluated exactly once, and the differences
between models are checked for significance rather than assumed.

The report closes by asking what these models would actually do at real-world celiac prevalence — which turns
out to matter more than any accuracy figure in it.

---

## Data Source

The dataset was obtained from Kaggle:
**Celiac Disease Lab Data** — <https://www.kaggle.com/datasets/jackwin07/celiac-disease-coeliac-disease>

To ensure balanced class representation, the dataset was shuffled and subsampled to include an equal number of
diagnosed and non-diagnosed individuals (363 each, n = 726). That processed file is committed as
`data/shuffled_data.csv`; the exact preparation steps are recorded in [`data/README.md`](data/README.md).

```{r load-data}
celiac_data <- read.csv("data/shuffled_data.csv")
head(celiac_data, 10)
```

## The features

- **PSS / DSS** — types of growth delay. PSS is proportionate, DSS is not.
- **IgA, IgG, IgM** — immunoglobulin antibodies in the blood. Elevated levels mean the immune system is engaged.
- **Marsh** — doctors use Marsh scores to quantify damage to the small intestine from celiac disease. The score
  reflects how much the intestinal villi have been damaged.
- **cd_type** — the clinical subtype of celiac disease.

```{r summary}
summary(celiac_data)
```

## Target Variable

The target `Disease_Diagnose` indicates whether a patient has been diagnosed with celiac disease. It is
converted to a factor for classification.

```{r target}
celiac_data$Disease_Diagnose <- as.factor(celiac_data$Disease_Diagnose)
table(celiac_data$Disease_Diagnose)
```

## Distribution of Numeric Features

Histograms help identify outliers and scale differences between features.

```{r histograms, fig.width=8, fig.height=5}
celiac_data %>%
  select(where(is.numeric)) %>%
  gather(attributes, value) %>%
  ggplot(aes(x = value)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 30) +
  facet_wrap(~attributes, scales = "free_x") +
  labs(x = "Values", y = "Frequency") +
  theme_bw()
```

Observations:

- Diabetes type is almost entirely type 1.
- There is little spread in the IgA values.
- Most patients are relatively young — roughly 1–15 years old.

## Clinical Subtypes of Celiac Disease

`cd_type` records the clinical subtype. Only diagnosed patients have a subtype, so `none` is excluded here.

```{r cd-type, fig.width=6, fig.height=4}
cd_type_only <- subset(celiac_data, cd_type != "none")

ggplot(cd_type_only, aes(x = cd_type)) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(x = "Subtype", y = "Count", title = "Count of each celiac subtype") +
  theme_bw()
```

```{r cd-type-pct}
round(prop.table(table(cd_type_only$cd_type)) * 100, 1)
```

## Feature Engineering

To prepare the data for modeling:

- Marsh scores were converted to numeric form (`3a` → `3.1`, `3b` → `3.2`, `3c` → `3.3`)
- Binary clinical symptoms were encoded as 0/1
- Gender and diagnosis were binarised
- Columns with extensive missing values or redundant information were excluded

```{r encode}
numeric_celiac_data <- celiac_data

numeric_celiac_data$Marsh <- gsub("a", ".1", numeric_celiac_data$Marsh)
numeric_celiac_data$Marsh <- gsub("b", ".2", numeric_celiac_data$Marsh)
numeric_celiac_data$Marsh <- gsub("c", ".3", numeric_celiac_data$Marsh)
numeric_celiac_data$Marsh <- as.numeric(numeric_celiac_data$Marsh)

numeric_celiac_data$Diabetes         <- ifelse(numeric_celiac_data$Diabetes == "Yes", 1, 0)
numeric_celiac_data$Abdominal        <- ifelse(numeric_celiac_data$Abdominal == "yes", 1, 0)
numeric_celiac_data$Sticky_Stool     <- ifelse(numeric_celiac_data$Sticky_Stool == "yes", 1, 0)
numeric_celiac_data$Weight_loss      <- ifelse(numeric_celiac_data$Weight_loss == "yes", 1, 0)
numeric_celiac_data$Gender           <- ifelse(numeric_celiac_data$Gender == "Female", 1, 0)
numeric_celiac_data$Disease_Diagnose <- ifelse(numeric_celiac_data$Disease_Diagnose == "yes", 1, 0)

head(numeric_celiac_data, 10)
```

### Choosing one predictor set

Every model below uses the same ten predictors. Three columns are dropped:

- **`cd_type`** — only diagnosed patients have a subtype, so it encodes the label directly.
- **`Diabetes.Type`** — `NA` for every non-diabetic patient and almost always `1` otherwise, so it adds no
  information beyond `Diabetes`.
- **`Diarrhoea`, `Short_Stature`** — multi-class categorical. KNN needs numeric input, and using them for the
  tree but not for KNN would make the model comparison unfair.

```{r feature-set}
PREDICTORS <- c("Age", "Gender", "Diabetes", "Abdominal", "Sticky_Stool",
                "Weight_loss", "IgA", "IgG", "IgM", "Marsh")

model_data <- numeric_celiac_data %>%
  select(all_of(PREDICTORS), Disease_Diagnose) %>%
  mutate(Disease_Diagnose = factor(Disease_Diagnose,
                                   levels = c(0, 1),
                                   labels = c("healthy", "sick")))

str(model_data)
```

## Feature Correlation Analysis

```{r correlation, fig.width=7, fig.height=7}
cor_data <- model_data %>%
  mutate(Disease_Diagnose = as.numeric(Disease_Diagnose) - 1) %>%
  cor()

corrplot(cor_data, method = "square")
```

As anticipated, most of the stomach-related symptoms are correlated with each other. The two features most
correlated with the diagnosis are **Marsh** and **Diabetes**. Abdominal pain, sticky stool and weight loss also
show correlation with the target.

The correlation between diabetes and the stomach symptoms prompted a check of the literature, to rule out that
these symptoms come from diabetes rather than from celiac disease.

> **Note on `Marsh`.** The Marsh score is the histological grading used to *make* the celiac diagnosis in the
> first place. Its strong correlation with the target is therefore partly circular. It is kept in the main
> models as the strongest available predictor, but a model fitted without it appears at the end of the report.

## Dimensionality Reduction (PCA)

PCA is used to visualise separation between healthy and sick individuals. The features are on very different
scales — `Age` spans decades while most symptoms are 0/1 — so the data is **scaled** before decomposition.
Without scaling, the first component would simply track `Age`.

```{r pca-variance}
pca_fit <- prcomp(model_data[, PREDICTORS], scale. = TRUE)

pct_var <- round(100 * summary(pca_fit)$importance[2, 1:2], 1)
pct_var
```

```{r pca-diagnosis, fig.width=6.5, fig.height=5}
pca_df <- data.frame(
  PC1 = pca_fit$x[, 1],
  PC2 = pca_fit$x[, 2],
  Diagnosis = model_data$Disease_Diagnose,
  Subtype   = celiac_data$cd_type
)

ggplot(pca_df, aes(x = PC1, y = PC2, colour = Diagnosis)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "PCA of celiac lab features",
    x = paste0("PC1 (", pct_var[1], "% variance)"),
    y = paste0("PC2 (", pct_var[2], "% variance)")
  ) +
  theme_bw()
```

```{r pca-subtype, fig.width=6.5, fig.height=5}
ggplot(pca_df, aes(x = PC1, y = PC2, colour = Subtype)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "PCA coloured by clinical subtype",
    x = paste0("PC1 (", pct_var[1], "% variance)"),
    y = paste0("PC2 (", pct_var[2], "% variance)")
  ) +
  theme_bw()
```

---

## One Shared Train/Test Split

The split is created **once** here and reused by all three models, so the comparison at the end is
apples-to-apples.

```{r split}
set.seed(1234)

n_total <- nrow(model_data)
train_idx <- sample(n_total, round(0.8 * n_total))

train_raw <- model_data[train_idx, ]
test_raw  <- model_data[-train_idx, ]

train_labels <- train_raw$Disease_Diagnose
test_labels  <- test_raw$Disease_Diagnose

# sanity check: the split must be disjoint and cover every row
stopifnot(
  length(intersect(train_idx, seq_len(n_total)[-train_idx])) == 0,
  nrow(train_raw) + nrow(test_raw) == n_total
)

c(train = nrow(train_raw), test = nrow(test_raw))
```

Both sets stay close to the 50/50 balance of the full dataset:

```{r split-balance}
rbind(
  train = prop.table(table(train_raw$Disease_Diagnose)),
  test  = prop.table(table(test_raw$Disease_Diagnose))
)
```

### Normalisation fitted on the training set only

KNN is distance-based, so features must be on a common scale. The min and max are computed from the
**training rows only** and then applied to the test rows — computing them over the full dataset would leak
information about the test set into the model.

Test values may therefore fall slightly outside `[0, 1]`; that is expected and correct.

```{r normalize}
train_min <- sapply(train_raw[, PREDICTORS], min)
train_max <- sapply(train_raw[, PREDICTORS], max)

apply_minmax <- function(df) {
  as.data.frame(scale(df[, PREDICTORS],
                      center = train_min,
                      scale  = train_max - train_min))
}

train_norm <- apply_minmax(train_raw)
test_norm  <- apply_minmax(test_raw)

range(test_norm)
```

Decision trees and random forests split on thresholds, so they are unaffected by any monotone rescaling. They
are fitted on the unnormalised features; only KNN uses `train_norm` / `test_norm`.

### A shared evaluation helper

Every model is scored with the same function, so the summary table at the end is consistent.

```{r eval-helper}
CLASSES <- c("healthy", "sick")

# forcing the levels keeps the matrix 2x2 even if a model never predicts one class
confusion <- function(truth, pred) {
  table(Predicted = factor(pred,  levels = CLASSES),
        Actual    = factor(truth, levels = CLASSES))
}

eval_model <- function(name, truth, pred) {
  cm <- confusion(truth, pred)
  tibble(
    Model             = name,
    Accuracy          = sum(diag(cm)) / sum(cm),
    Sensitivity       = cm["sick", "sick"] / sum(cm[, "sick"]),
    Specificity       = cm["healthy", "healthy"] / sum(cm[, "healthy"]),
    `False negatives` = cm["healthy", "sick"],
    `False positives` = cm["sick", "healthy"]
  )
}
```

---

## Model Selection by Cross-Validation

The hyperparameters below — `k` for KNN, `mtry` for the forest, and the false-negative cost ratio for C5.0 —
were originally chosen by hand. Here they are selected properly, by **repeated 10-fold cross-validation on the
training set only**. The test set is not touched until the final comparison.

```{r cv-folds}
set.seed(2024)
folds <- vfold_cv(train_raw, v = 10, repeats = 5, strata = Disease_Diagnose)

nrow(folds)  # 10 folds x 5 repeats
```

Each candidate is scored with the same loop. Note that the min-max scaler for KNN is refitted **inside every
fold** from that fold's analysis rows — doing it once outside would leak the assessment rows' range into
training, which is the bug this report fixed earlier. Putting it inside the loop makes that mistake structurally
impossible to reintroduce.

```{r cv-helper}
cv_evaluate <- function(fit_predict) {
  map_dfr(seq_len(nrow(folds)), function(i) {
    sp <- folds$splits[[i]]
    tr <- analysis(sp)
    te <- assessment(sp)

    cm <- confusion(te$Disease_Diagnose, fit_predict(tr, te))

    tibble(
      accuracy = sum(diag(cm)) / sum(cm),
      sens     = cm["sick", "sick"] / sum(cm[, "sick"]),
      fn       = cm["healthy", "sick"]
    )
  })
}

# mean and standard error across the 50 resamples
cv_summarise <- function(res, label) {
  res %>% summarise(
    Setting     = label,
    Accuracy    = mean(accuracy),
    `Std error` = sd(accuracy) / sqrt(n()),
    Sensitivity = mean(sens),
    `Mean FN`   = mean(fn)
  )
}
```

### Tuning `k` for KNN

```{r cv-knn, cache=TRUE}
k_grid <- seq(1, 25, by = 2)

knn_cv <- map_dfr(k_grid, function(k) {
  cv_evaluate(function(tr, te) {
    lo <- sapply(tr[, PREDICTORS], min)
    hi <- sapply(tr[, PREDICTORS], max)
    rng <- ifelse(hi - lo == 0, 1, hi - lo)     # guard against constant columns

    tr_n <- scale(tr[, PREDICTORS], center = lo, scale = rng)
    te_n <- scale(te[, PREDICTORS], center = lo, scale = rng)

    knn(train = tr_n, test = te_n, cl = tr$Disease_Diagnose, k = k)
  }) %>% cv_summarise(k)
})

best_k <- knn_cv$Setting[which.max(knn_cv$Accuracy)]
knn_cv
```

Two things are worth noting rather than glossing over. First, `k = 1` winning is a mild warning sign: it usually
means the classes are close to separable, which here is the `Marsh` effect again rather than a genuinely
excellent nearest-neighbour structure. Second, the curve is flat — the spread across all values of k is only
about two standard errors, so this choice is far less decisive than picking the maximum suggests.

There is also a tension in the selection criterion: `k = 1` maximises accuracy, but `k = 9` produces noticeably
fewer false negatives (1.18 vs 1.74 per fold). Since this project treats false negatives as the more costly
error, `k = 9` is arguably the better choice. Accuracy is kept as the criterion here because selecting purely on
false negatives is degenerate in general — a model that predicts "sick" for everyone scores zero — but the
disagreement is real and is revisited in the limitations.

```{r cv-knn-plot, fig.width=6.5, fig.height=4}
ggplot(knn_cv, aes(x = Setting, y = Accuracy)) +
  geom_ribbon(aes(ymin = Accuracy - 1.96 * `Std error`,
                  ymax = Accuracy + 1.96 * `Std error`), alpha = 0.2) +
  geom_line() + geom_point() +
  geom_vline(xintercept = best_k, linetype = "dashed", colour = "firebrick") +
  labs(title = paste0("KNN: cross-validated accuracy by k (best = ", best_k, ")"),
       x = "k", y = "CV accuracy ± 1.96 SE") +
  theme_bw()
```

### Tuning `mtry` for the random forest

```{r cv-rf, cache=TRUE}
mtry_grid <- 1:6

rf_cv <- map_dfr(mtry_grid, function(m) {
  cv_evaluate(function(tr, te) {
    fit <- ranger::ranger(Disease_Diagnose ~ ., data = tr[, c(PREDICTORS, "Disease_Diagnose")],
                          num.trees = 500, mtry = m, seed = 234)
    predict(fit, te)$predictions
  }) %>% cv_summarise(m)
})

best_mtry <- rf_cv$Setting[which.max(rf_cv$Accuracy)]
rf_cv
```

### Tuning the C5.0 false-negative cost ratio

Accuracy is the wrong objective here — the point of the cost matrix is to trade false positives for false
negatives, so the ratio is chosen on **mean false negatives per fold**, with accuracy reported alongside to show
what that costs.

```{r cv-cost, cache=TRUE}
cost_grid <- c(1, 2, 3, 5, 10)

cost_cv <- map_dfr(cost_grid, function(r) {
  cv_evaluate(function(tr, te) {
    costs <- matrix(c(0, r, 1, 0), nrow = 2,
                    dimnames = list(actual = CLASSES, predicted = CLASSES))
    fit <- C5.0(tr[, PREDICTORS], tr$Disease_Diagnose, trials = 10, costs = costs)
    predict(fit, te)
  }) %>% cv_summarise(r)
})

best_cost <- cost_cv$Setting[which.min(cost_cv$`Mean FN`)]
cost_cv
```

The trade-off is visible in the table: false negatives fall steadily as the ratio rises, while accuracy peaks
around 3 and then declines. Selecting on false negatives picks the most aggressive ratio and knowingly gives up
about a point of accuracy — which is the entire point of using a cost matrix.

```{r cv-selected}
tibble(
  Parameter = c("KNN k", "Forest mtry", "C5.0 cost ratio"),
  Selected  = c(best_k, best_mtry, best_cost)
)
```

These selected values are used for every model fitted below.

---

## Decision Tree (C5.0)

```{r tree-fit}
celiac_model <- C5.0(train_raw[, PREDICTORS], train_labels)
celiac_prediction <- predict(celiac_model, test_raw)

CrossTable(test_labels, celiac_prediction,
           prop.chisq = FALSE, prop.c = FALSE, prop.r = FALSE,
           dnn = c("actual", "predicted"))
```

### Boosting

```{r tree-boost}
celiac_boost10 <- C5.0(train_raw[, PREDICTORS], train_labels,
                       trials = 10, control = C5.0Control(CF = 0.2))

celiac_boost_pred10 <- predict(celiac_boost10, test_raw)

CrossTable(test_labels, celiac_boost_pred10,
           prop.chisq = FALSE, prop.c = FALSE, prop.r = FALSE,
           dnn = c("actual", "predicted"))
```

```{r tree-plot, fig.width=20, fig.height=10}
plot(celiac_boost10, type = "simple", trial = 0)
```

### Adding a cost matrix

Clinically, a false negative (predicting healthy when the patient is sick) is more dangerous than a false
positive, so it is given three times the cost.

**Watch the orientation.** `C5.0()` expects the cost matrix indexed as `[actual, predicted]` — rows are the
true class, columns are the predicted one. Several textbook examples label the dimensions the other way round,
and since the labels are only cosmetic, a transposed matrix fails silently: it penalises false *positives* while
appearing to penalise false negatives. Here that would push false negatives up rather than down.

The false-negative cell is therefore *actual = sick, predicted = healthy*, i.e. row `sick`, column `healthy`.

```{r cost-matrix}
matrix_dimensions <- list(actual = CLASSES, predicted = CLASSES)

# column-major fill puts best_cost at [sick, healthy] = actual sick, predicted healthy = false negative
error_cost <- matrix(c(0, best_cost, 1, 0), nrow = 2, dimnames = matrix_dimensions)
error_cost
```

```{r tree-cost}
celiac_cost_10 <- C5.0(train_raw[, PREDICTORS], train_labels,
                       trials = 10, costs = error_cost)

celiac_cost_pred <- predict(celiac_cost_10, test_raw)

CrossTable(test_labels, celiac_cost_pred,
           prop.chisq = FALSE, prop.c = FALSE, prop.r = FALSE,
           dnn = c("actual", "predicted"))
```

The penalty works: false negatives fall from 5 to 2 on the test set, and — unusually for a cost-sensitive
model — overall accuracy *improves* rather than trading off. The remaining errors are mostly false positives,
which in this setting means recommending a confirmatory test for a patient who turns out to be healthy: the
much cheaper mistake.

A note on what cross-validation changed here. An earlier version of this report hand-picked a 3:1 ratio and got
**zero** false negatives on this test set. That looked like the headline result, but it did not survive proper
tuning: cross-validation selects a 10:1 ratio, and the honest test-set score is 2 false negatives, not 0. The
perfect result was partly luck in a single 145-patient split — exactly the kind of overstatement that repeated
resampling exists to catch.

```{r tree-cost-plot, fig.width=20, fig.height=10}
plot(celiac_cost_10, type = "simple", trial = 0)
```

---

## K-Nearest Neighbors

KNN is a simple distance-based baseline, fitted on the normalised features using the cross-validated value of
k selected above.

```{r knn}
knn_pred <- knn(train = train_norm, test = test_norm, cl = train_labels, k = best_k)

CrossTable(x = test_labels, y = knn_pred, prop.chisq = FALSE,
           dnn = c("actual", "predicted"))
```

---

## Random Forest

A random forest should reduce the overfitting of a single tree and generalise better, and gives feature
importance for free.

```{r rf-fit}
rf_train <- train_raw   # already contains Disease_Diagnose alongside the predictors

rf_mod <-
  rand_forest(trees = 1000, mtry = best_mtry) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

set.seed(234)
rf_fit <- rf_mod %>% fit(Disease_Diagnose ~ ., data = rf_train)
```

```{r rf-predict}
rf_testing_pred <-
  predict(rf_fit, test_raw) %>%
  bind_cols(predict(rf_fit, test_raw, type = "prob")) %>%
  bind_cols(Label = test_labels)

rf_testing_pred %>% accuracy(truth = Label, .pred_class)
```

### ROC curve

`Label` has levels `healthy`, `sick`. `yardstick` treats the **first** level as the event by default, so the
curve is drawn against `.pred_healthy`.

```{r rf-roc, fig.width=5.5, fig.height=5}
rf_testing_pred %>%
  roc_curve(truth = Label, .pred_healthy) %>%
  autoplot()
```

```{r rf-auc}
rf_testing_pred %>% roc_auc(truth = Label, .pred_healthy)
```

### Feature importance

```{r rf-vip, fig.width=6.5, fig.height=4.5}
vip(rf_fit, num_features = 10) +
  labs(title = "Random forest feature importance") +
  theme_bw()
```

`Marsh` dominates — which is exactly the circularity flagged earlier. The next section quantifies how much of
the performance depends on it.

---

## How Much of the Performance is `Marsh`?

The Marsh score is the histological criterion clinicians use to diagnose celiac disease. A model built on it is
close to predicting the label from the label. To see what the other features contribute on their own, the same
random forest is refitted with `Marsh` removed.

```{r rf-no-marsh}
train_no_marsh <- train_raw %>% select(-Marsh)
test_no_marsh  <- test_raw  %>% select(-Marsh)

set.seed(234)
rf_fit_no_marsh <- rf_mod %>% fit(Disease_Diagnose ~ ., data = train_no_marsh)

rf_no_marsh_pred <- predict(rf_fit_no_marsh, test_no_marsh)$.pred_class

eval_model("Random forest (no Marsh)", test_labels, rf_no_marsh_pred)
```

```{r rf-no-marsh-vip, fig.width=6.5, fig.height=4.5}
vip(rf_fit_no_marsh, num_features = 9) +
  labs(title = "Feature importance without Marsh") +
  theme_bw()
```

---

## Model Comparison

All rows below are computed on the **same 145 test patients** using the **same ten predictors**.

```{r comparison}
comparison <- bind_rows(
  eval_model("Decision tree (C5.0)",           test_labels, celiac_prediction),
  eval_model("C5.0 + boosting (10 trials)",    test_labels, celiac_boost_pred10),
  eval_model("C5.0 + boosting + cost matrix",  test_labels, celiac_cost_pred),
  eval_model(paste0("KNN (k = ", best_k, ")"), test_labels, knn_pred),
  eval_model("Random forest (1000 trees)",     test_labels, rf_testing_pred$.pred_class),
  eval_model("Random forest (no Marsh)",       test_labels, rf_no_marsh_pred)
)

comparison %>% mutate(across(c(Accuracy, Sensitivity), ~ round(.x, 3)))
```

```{r comparison-plot, fig.width=7, fig.height=4}
comparison %>%
  mutate(Model = factor(Model, levels = rev(Model))) %>%
  ggplot(aes(x = `False negatives`, y = Model)) +
  geom_col(fill = "lightblue", color = "black") +
  labs(title = "False negatives by model (lower is better)",
       x = "False negatives on the test set", y = NULL) +
  theme_bw()
```

### Are these differences real?

A gap of one or two patients out of 145 is easy to over-read. **McNemar's test** compares two classifiers on the
*same* test cases by looking only at the cases where they disagree, which is the appropriate test for paired
predictions.

```{r mcnemar}
mcnemar_compare <- function(name_a, pred_a, name_b, pred_b) {
  tab <- table(A = pred_a == test_labels, B = pred_b == test_labels)
  p   <- mcnemar.test(tab)$p.value
  tibble(
    Comparison   = paste(name_a, "vs", name_b),
    `A only correct` = tab["TRUE", "FALSE"],
    `B only correct` = tab["FALSE", "TRUE"],
    `p value`    = round(p, 3)
  )
}

bind_rows(
  mcnemar_compare("Cost-sensitive C5.0", celiac_cost_pred, "Random forest", rf_testing_pred$.pred_class),
  mcnemar_compare("Cost-sensitive C5.0", celiac_cost_pred, "Decision tree",  celiac_prediction),
  mcnemar_compare("Random forest",       rf_testing_pred$.pred_class, "KNN", knn_pred)
)
```

The discordant counts are small, so these comparisons have very little statistical power. Read the p-values as
*"this test set cannot distinguish these models"* rather than as evidence they perform equally — with 145
patients, only a large difference would reach significance. The cross-validated estimates earlier in the report,
averaged over 50 resamples, are the more reliable basis for choosing between them.

---

## What Happens at Realistic Prevalence

Every accuracy figure above was measured on a dataset that is **50% celiac by construction**. Real celiac
prevalence in the general population is roughly **1%** (Singh et al., *Clin Gastroenterol Hepatol* 2018, pooled
seroprevalence ~1.4%, biopsy-confirmed ~0.7%). That gap matters more than it first appears.

Sensitivity and specificity are properties of the classifier and do not depend on prevalence. **Precision — the
positive predictive value — does.** So the models can be projected onto a realistic population directly from
their sensitivity and specificity, without refitting anything:

$$\text{PPV} = \frac{\text{sens} \times p}{\text{sens} \times p + (1-\text{spec}) \times (1-p)}
\qquad
\text{NPV} = \frac{\text{spec} \times (1-p)}{\text{spec} \times (1-p) + (1-\text{sens}) \times p}$$

```{r prevalence-helpers}
ppv_at <- function(sens, spec, p) (sens * p) / (sens * p + (1 - spec) * (1 - p))
npv_at <- function(sens, spec, p) (spec * (1 - p)) / (spec * (1 - p) + (1 - sens) * p)
```

Sanity check — at the dataset's own prevalence the formula must reproduce the observed precision:

```{r prevalence-check}
best     <- comparison %>% filter(Model == "C5.0 + boosting + cost matrix")
cm_best  <- confusion(test_labels, celiac_cost_pred)
obs_prev <- mean(test_labels == "sick")

round(c(
  from_formula = ppv_at(best$Sensitivity, best$Specificity, obs_prev),
  from_matrix  = cm_best["sick", "sick"] / sum(cm_best["sick", ])
), 4)
```

### Predictive value across prevalence

```{r prevalence-table}
prevalences <- c(0.005, 0.01, 0.05, 0.50)

comparison %>%
  filter(Model %in% c("C5.0 + boosting + cost matrix", "Random forest (1000 trees)")) %>%
  rowwise() %>%
  reframe(
    Model, Prevalence = prevalences,
    PPV = ppv_at(Sensitivity, Specificity, prevalences),
    NPV = npv_at(Sensitivity, Specificity, prevalences)
  ) %>%
  mutate(across(c(PPV, NPV), ~ round(.x, 3)),
         Prevalence = scales::percent(Prevalence, accuracy = 0.1))
```

```{r prevalence-curve, fig.width=7, fig.height=4.5}
curve_df <- comparison %>%
  filter(Model %in% c("C5.0 + boosting + cost matrix", "Random forest (1000 trees)")) %>%
  rowwise() %>%
  reframe(Model, p = 10^seq(-3, log10(0.5), length.out = 200)) %>%
  left_join(comparison %>% select(Model, Sensitivity, Specificity), by = "Model") %>%
  mutate(PPV = ppv_at(Sensitivity, Specificity, p))

ggplot(curve_df, aes(x = p, y = PPV, colour = Model)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 0.01, linetype = "dashed", colour = "firebrick") +
  geom_vline(xintercept = 0.50, linetype = "dotted") +
  annotate("text", x = 0.011, y = 0.05, label = "real prevalence ~1%",
           hjust = 0, size = 3.2, colour = "firebrick") +
  annotate("text", x = 0.5, y = 0.05, label = "this dataset (50%)",
           hjust = 1.05, size = 3.2) +
  scale_x_log10(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(title = "Precision collapses at realistic prevalence",
       x = "Disease prevalence (log scale)", y = "Positive predictive value") +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Why accuracy is the wrong headline

At 1% prevalence, a model that simply predicts "healthy" for everyone achieves **99% accuracy** — better than
anything reported above — while finding zero cases. Accuracy on a balanced sample says almost nothing about
performance as a screening tool.

The honest reading of these models is:

- **NPV stays near 100%.** A negative result is highly trustworthy.
- **PPV falls sharply.** At 1% prevalence most positive flags would be false alarms.

That combination describes a **rule-out screening test**, not a diagnostic one: useful for deciding who does
*not* need a biopsy, not for confirming who has celiac disease. Given that biopsy is the confirmatory step
anyway, that is a reasonable role — but it is a much narrower claim than
"`r scales::percent(best$Accuracy, accuracy = 0.1)` accurate".

---

## Conclusions

- **The cost-sensitive boosted tree and tuned KNN tie for best**, both at 97.2% accuracy with 2 false negatives
  and 2 false positives. Encoding the clinical asymmetry into C5.0 training works — it takes false negatives
  from 5 to 2 while *improving* accuracy.
- **No difference between the top models is statistically significant.** McNemar's test on the paired
  predictions gives p ≈ 0.07 at best. With 145 patients, this test set cannot separate them; the
  cross-validated estimates over 50 resamples are the more trustworthy comparison.
- The random forest did **not** come out ahead (93.8%, 7 false negatives). An earlier version of this analysis
  concluded that it did — an artefact of scoring each model on a different train/test split and a different
  feature set. Once the comparison is fair, the advantage disappears.
- **The headline accuracy is inflated by `Marsh`.** Dropping it costs about seven points of accuracy and
  triples the false positives, showing how much apparent performance comes from a feature that is essentially
  part of the diagnosis itself.
- **At realistic prevalence these models are rule-out tools, not diagnostics.** Precision falls from ~97% on
  this balanced sample to roughly **23% at 1% prevalence**, while NPV stays at ~100%. A negative result is
  trustworthy; a positive one mostly is not.

Several of these conclusions reverse what the first version of this report claimed, and every reversal came
from fixing methodology rather than changing a model: sharing one split and feature set, correcting the
orientation of the cost matrix, tuning by cross-validation instead of by hand, and testing whether the
differences were real. Notably, the earlier "zero false negatives" result did not survive proper tuning.

## Limitations

- **Target leakage.** `Marsh` is the histological grading used to make the diagnosis, so the main models are
  partly predicting the label from the label.
- **A small test set.** Hyperparameters are now chosen by repeated cross-validation, but the final comparison
  still rests on 145 patients, and McNemar's test confirms it cannot separate the leading models. Treat the
  ranking as provisional.
- **Artificially balanced sample.** n = 726, subsampled to 50/50, so headline accuracy does not transfer to a
  screening population — see the prevalence section for what it becomes at a realistic base rate.
- **Selection criteria are not fully consistent.** `k` and `mtry` are chosen on accuracy while the cost ratio is
  chosen on false negatives. For KNN specifically, `k = 9` would roughly halve false negatives at a cost of one
  accuracy point; a single, explicitly clinical objective would be more coherent.
- **Tuning and final evaluation share a training set.** The split is honest — the test set is touched once — but
  a fully nested cross-validation would give a less optimistic estimate still.
- **No external validation** on an independent cohort.
- This is a learning exercise. The models are not suitable for clinical use.

## Next Steps

- Nested cross-validation, so the performance estimate is not conditioned on a single outer split
- A single explicit objective (e.g. expected cost, or sensitivity at fixed specificity) used to select every
  hyperparameter, resolving the inconsistency noted above
- Decision-threshold analysis on predicted probabilities, replacing the discrete cost ratio with a full
  precision–recall trade-off curve
- One-hot encode `Diarrhoea` and `Short_Stature` so they can be used without breaking KNN
- Investigate whether different symptoms correlate with different celiac subtypes
- Unsupervised clustering, to see whether the data splits into more than two groups without the labels
- External validation on an independent dataset
